---
title: "Linking delay with Wx and NOTAMs"
author: "Dan Flynn"
output: html_document
---


Delay data: Processed by `Delay_data.R`, which produces `flightmeasures5c.RData`

```{r setup, echo = F, message=F}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(readxl)
library(gridExtra)
library(sjPlot)    # For nicer output tables of the statistical models
library(plotly)    # for interactive plots
library(kableExtra)# for better tables

# Working directory: This sets the path to my H: drive if being run from my computer, otherwise put in your own path
workingdir <- ifelse(grepl('flynn', system('whoami', intern = T)),
                     "~/NOTAM_Local/",
                     "<Put_Your_Path_Here>")

setwd(workingdir)
knitr::opts_knit$set(root.dir = workingdir)

# Load delay data
load("flightmeasures5c.RData")

# Load METAR
metar <- read.csv('METAR_extract.csv',
                  na.strings = 'M')

# Load NOTAMs
notam <- read_xlsx('notam_extract.xlsx', sheet = 'data')

```

Date ranges:
METARS from 2017-04-26 to 2019-06-29, with 765,764 records. 

```{r metar_summary}
m = metar %>%
  mutate(vsby = as.numeric(vsby),
         skyl1 = as.numeric(skyl1),
         skyl2 = as.numeric(skyl2),
         skyl3 = as.numeric(skyl3),
         skyl4 = as.numeric(skyl4),
         valid = as.POSIXct(as.character(valid), format = "%Y-%m-%d %H:%M", tz = "UTC"),
         date = format(valid, "%Y-%m-%d"),
         hr = format(valid, "%Y-%m-%d"))

m %>%
  group_by(station) %>%
  summarize(vis_min = min(vsby, na.rm=T),
            vis_max = max(vsby, na.rm=T),
            vis_mean = mean(vsby, na.rm=T))

range(m$valid, na.rm = T)
```

Match the NOTAM data to these METARs. NOTAM data are for the similar time range, from April 2017 to June 2019, with 6,717 records in total. After matching to the same stations as the METARS, there are 2,142 NOTAM records.

```{r match_notam_metar}
n = notam %>%
  filter(notam_location %in% ifelse(nchar(as.character(metar$station)) == 3, paste0("K", metar$station), metar$station)) %>%
  mutate(station = as.factor(notam_location),
         start_time = as.POSIXct(start_time, format = "%Y-%m-%d %H:%M", tz = "UTC"),
         end_time = as.POSIXct(end_time, format = "%Y-%m-%d %H:%M", tz = "UTC"),
         start_hr = format(start_time, format = "%Y-%m-%d %H"),
         end_hr = format(end_time, format = "%Y-%m-%d %H"))

# Count active by station, date, and hour
# Simple way: loop, place in date range by hour
hr_range = with(n, range(start_time, end_time))
start_day = format(min(hr_range), '%Y-%m-%d')
end_day = format(max(hr_range), '%Y-%m-%d')

day_hour <- seq(from = as.POSIXct(paste0(start_day, "00:00"), tz = 'UTC'), 
                  to = as.POSIXct(paste0(end_day, " 23:00"), tz = 'UTC'),
                  by = "hour")

day_hr_ct <- vector()

for(i in day_hour){
  # i = day_hour[100]
  n_filt = n %>%
    filter(start_time <= i & end_time >= i) %>%
    group_by(notam_location) %>%
    summarize(count_active = n()) %>%
    mutate(day_hr = i)
  
  day_hr_ct = rbind(day_hr_ct, n_filt)

  }

day_hr_ct = day_hr_ct %>%
  mutate(day_hr = as.POSIXct(day_hr, origin = '1970-01-01', tz = 'UTC'))

head(day_hr_ct)
```

Delay data: 83,218 flights in this subset.

```{r match_notam_metar_flights}
# First, convert all time variables to date-time in UTC

timevars = names(d)[grep('_time_', names(d))]
# Omit 'priority' variables
timevars = timevars[!timevars %in% timevars[grep('priority', timevars)]]

# Omit any variables which are all NA
timevars = timevars[!apply(d[,timevars], 2, function(x) all(is.na(x)))]

use_vars = c('airspaceuser', 'callsign',
             'destination', 'origin', 'distance', 
             timevars,
             'delay')
dx <- d %>%
  select(use_vars) %>%
  mutate_at(timevars, .funs = function(x) as.POSIXct(x, tz = 'UTC', origin = '1970-01-01'))



```


## Plotting

Plot a dual-axis figure for visibility and sky level 1

```{r}
m_filter = m %>%
  filter(station %in% c("MSY", "PHL", "LAX"
                        )) %>%
  filter(date >= '2019-04-01' & date < '2019-05-01')

sky_plot = ggplot(m_filter, aes(group = station)) +
  geom_line(aes(x = valid, y = skyl1), color = 'blue') +
  facet_wrap(~station) +
  xlab("Date") +
  ylab("Sky Level 1 Altitude (ft)") +
  theme_bw()

vis_plot = ggplot(m_filter, aes(group = station)) +
  geom_line(aes(x = valid, y = vsby), color = 'red') +
  facet_wrap(~station) +
  xlab("Date") +
  ylab("Visibility (mi)") +
  theme_bw()

grid.arrange(sky_plot, vis_plot)
```

```{r}
# Zoom to one week, one location. Put vis in feet, 5280' per mile
m_filter2 = m_filter %>% 
  mutate(vis_ft = vsby*5280) %>% 
  filter(date >= "2019-04-02" & date <= "2019-04-09" & station == 'MSY')

sky_plot = ggplot(m_filter2) +
  geom_line(aes(x = valid, y = skyl1), color = 'blue') +
  xlab("Date") +
  ylab("Sky Level 1 Altitude (ft) \n Limit 5000'") +
  ylim(0, 5000) +
  theme_bw() +
  ggtitle('MSY METAR sample, 2019-04-02 to 2019-04-09')

vis_plot = ggplot(m_filter2) +
  geom_line(aes(x = valid, y = vis_ft), color = 'red') +
  xlab("Date") +
  ylab("Visibility (ft) \n Limit 10,000'") +
  ylim(0, 10000) +
  theme_bw()

grid.arrange(sky_plot, vis_plot)
# Save to file with arrangeGrob
g <- arrangeGrob(sky_plot, vis_plot)
g
ggsave('METAR_example_plot_MSY.jpeg', g)
```

### Plot active NOTAMs
First, simple plot of count by station and active date range

```{r}
ggplot(day_hr_ct,
       aes(x = day_hr,
           y = count_active,
           group = notam_location)) +
  geom_line() +
  facet_wrap(~notam_location)


# Just plot MSY
NotFancy <- function(l) {
  l <- formatC(round(l, 0), width = 6, flag = " ")
  parse(text=l)
}

notam_count_plot <- ggplot(day_hr_ct %>% 
         filter(notam_location == "KMSY" &
                  day_hr >= '2019-04-02' & day_hr <= '2019-04-09'),
       aes(x = day_hr,
           y = count_active,
           group = notam_location)) +
  geom_line(color = 'darkgreen', size = 1) +
  scale_y_continuous(name = "Count of active messages",
                     breaks = seq(0, 100, 1),
                     labels = NotFancy)+#scales::comma) +
  theme_bw() +
  ggtitle('Count of active NOTAMs in MSY') 

notam_count_plot

grid.arrange(notam_count_plot, sky_plot, vis_plot)
# Save to file with arrangeGrob
g <- arrangeGrob(notam_count_plot, sky_plot, vis_plot)

ggsave('METAR_example_plot_MSY_plus_NOTAMS.jpeg', g)
```

